\chapter{Sketch Proof of Decidable Type-Checking}
\label{sec:decproof}

This is a proof sketch that the type system of \ATAL is decidable. It follows
the structure of the file \texttt{Lemmas/TermDec.agda}, which contains our
Agda-checked proof. The main result of the file is the following function:

\begin{code}
\>\AgdaFunction{programstate-dec} \AgdaSymbol{:} \AgdaSymbol{∀} \AgdaBound{P} \AgdaSymbol{→} \AgdaDatatype{Dec} \AgdaSymbol{(}\AgdaDatatype{⊢} \AgdaBound{P} \AgdaDatatype{programstate}\AgdaSymbol{)}\<%
\end{code}

We will here present a slightly simplified sketch of how the result is shown in
that file. The actual proof bodies are quite repetitive and have been left
out. Instead we only present the structure of the proofs, which is much more
interesting.

\begin{lemma}
  \label{lemma:dec-v}
  For any $\Psi_{\mathrm{g}},\Delta, \high v,\Gamma$ the following holds:

  \begin{itemize}
  \item When given derivations of
    $\ofvval{\Psi_{\mathrm{g}},\Delta}{\high v}{\Gamma => \tau}$ and
    $\ofvval{\Psi_{\mathrm{g}},\Delta}{\high v}{\Gamma => \tau'}$ we can
    conclude that $\tau = \tau'$.
  \item We decide if any $\tau$ exists such that
    $\ofvval{\Psi_{\mathrm{g}},\Delta}{\high v}{\Gamma => \tau}$ is derivable.
  \end{itemize}

  We can prove similar lemmas for instructions and instruction sequences.
\end{lemma}


\begin{lemma}
  \label{lemma:dec-g}
  For any $\high g$ the following holds:

  \begin{itemize}
  \item We can find a $\tau$ such that any derivation of
    $\ofgval{\Psi_{\mathrm{g}}}{\high g}{\tau'}$ will have $\tau = \tau'$.
  \item When given $\Psi_{\mathrm{g}}$ and $\tau$ we can decide if
    $\ofgval{\Psi_{\mathrm{g}}}{\high g}{\tau}$ is derivable.
  \end{itemize}

  We can prove a similar lemma for global collections.
\end{lemma}

\begin{lemma}
  For any $\Psi_{\mathrm{g}},\Psi_{\mathrm{h}},\high w$ one of the following
  is true:

  \begin{itemize}
  \item No $\tau$ exists making
    $\ofwval{\Psi_{\mathrm{g}},\Psi_{\mathrm{h}}}{\high w}{\tau}$ derivable or
  \item We can find a $\tau$ such that:
    \begin{itemize}
    \item $\ofwval{\Psi_{\mathrm{g}},\Psi_{\mathrm{h}}}{\high w}{\tau}$ is
      derivable and
    \item When given $\tau'$ we can decide if
      $\ofwval{\Psi_{\mathrm{g}},\Psi_{\mathrm{h}}}{\high w}{\tau'}$ is
      derivable. If it is derivable, then $\subtype{\nil}{\tau}{\tau'}$ is also
      derivable.
    \end{itemize}
  \end{itemize}

  We can prove similar lemmas for stacks and register files.
\end{lemma}

\begin{lemma}
  For any $\high h$ the following holds:

  \begin{itemize}
  \item We can find a $\tau$ such that any derivation of
    $\ofhval{\Psi_{\mathrm{g}},\Psi_{\mathrm{h}}}{\high h}{\tau'}$ will have
    $\subtype{\nil}{\tau}{\tau'}$.
  \item When given a $\Psi_{\mathrm{g}},\Psi_{\mathrm{h}}$ and $\tau$ we can
    decide if $\ofhval{\Psi_{\mathrm{g}}}{\high h}{\tau}$ is derivable.
  \end{itemize}

  We can prove a similar lemma for heap collections.
\end{lemma}

\begin{lemma}
  For any $\Psi_{\mathrm{g}},\high H, \high R, \high I$ the following is true:

  \begin{itemize}
  \item We can decide if any $\Psi_{\mathrm{h}},\Gamma$ exists such that
    $\ofprogramstate{\Psi_{\mathrm{g}}}{(\high H, \high R, \high
      I)}{(\Psi_{\mathrm{h}},\Gamma)}$ is derivable.
  \end{itemize}
\end{lemma}

\begin{lemma}
  For any $\high P$, we can decide if $\valid{}{P}$ is derivable.
\end{lemma}
