\input{preamble}
\input{macros}

\begin{document}

\begin{titlepage}
\maketitle
\end{titlepage}
\tableofcontents

\input{abstract}

\chapter{Introduction}
\input{introduction/motivation}
\input{introduction/proof-carrying-code}
\input{introduction/problem-statement}

\chapter{Typed Assembly Language}
This chapter has three parts. First we will formally introduce the concepts of
Typed Assembly Language in an informal setting. We will then give a formal
definition of the same concepts. Finally we will introduce the specific instance
of Typed Assembly Language used in \cite{STAL}.

Note that I have chosen a specific way of presenting Typed Assembly Language,
that lends itself well to formalization. While this approach is not completely
similar to the approach used in \cite{STAL}, I expect that the concepts should
be fairly recognizable.

\input{tal/informal}
\input{tal/formal}
\input{tal/stal}

\chapter{Stack-Based Typed Assembly Language}
This chapter will define the two languages $STAL$ and $STAL_S$. The design bla
bla previous section.

For the most part $STAL$ is equivalent to the one defined in \cite{STAL},
however a few changes have been made to simplify the proofs. It is our opinion
that the resulting language is morally equivalent to the original one.

\input{stal/simple}
\input{stal/typed}

\chapter{Lessons Learned: Tips for doing formalizations in Agda}
\input{lessons/deceq}
\input{lessons/eqhelp}
\input{lessons/substitutions}

% \chapter{Judgments}
% \input{judgments/grammar}
% \input{judgments/types}
% \input{judgments/subtypes}
% \input{judgments/substitution}
% \input{judgments/run}
% \input{judgments/terms}
% \input{judgments/semantics}

% \chapter{Lemmas}
% \input{lemmas/misc}
% \input{lemmas/eq}
% \input{lemmas/types}
% \input{lemmas/subtypes}
% \input{lemmas/substitution}
% \input{lemmas/run}

\listoftodos[Notes]

\bibliographystyle{plain}
\bibliography{report}
\end{document}
